Doublet splitting and linear amplitude characteristics of fusion alpha-particle driven ion cyclotron emission

### 2022-02-02 manuscript.tex
tiny word change in caption
updates to calc of pitch -0.646

Solutions to the linearised Maxwell-Vlasov system of equations are found in $(k_\parallel, k_\bot)$ space for a ring-beam distribution of energetic minority alpha-particles and Maxwellian electrons and background thermal majority deuteron ions. Calculations are made for alpha-particles with ring-beams of two pitch-angle cosine values: $u_\parallel=-0.64 u_0 \simeq -1.085 V_A$, which is associated with the ion-cyclotron emission observed on JET Pulse No. 26148, see Figs. 3 and 15 and associated text in Ref. \cite{Cottrell1993}; and $u_\parallel=0 u_0$ representing a ring distribution with no parallel drift, see Fig. 1(a) of Ref \cite{Cook2013}. Here $u_0=\sqrt{u_\parallel^2 + u_\perp^2}$ is the speed of an alpha-particle with an energy of 3.6 MeV. The complex frequency solutions, $\omega$, are found corresponding to the fast Alfv{\'e}n branch on a grid of $(k_\parallel, k_\bot)$.

The plasma parameters approximate those found in the region indicated in Fig. 3 of Ref. \cite{Cottrell1993}, which are associated with the continuous trace signal of Fig. 2 (ibid). The electrons and deuterons, both modelled as Maxwellians, have temperatures of 1 keV and the electrons have a number density of $1.7\times 10^{19}$ $\mathrm{m^{-3}}$. The magnetic field is $2.07$ $\mathrm{T}$.
The ring-beam distribution function, Eq. \ref{eq:alphaf}, is a separable function of $v_\parallel$, i.e. the beam given by Eq. 1, and $v_\perp$, i.e. the ring given by Eq. 2. The component parallel (perpendicular) to the magnetic field is a drifting Maxwellian with thermal speed $v_{\parallel, th}$ ($v_{\perp, th}$) and drift velocity $u_\parallel$ ($u_\perp$). The thermal spread of the alpha particles' distribution function $v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The concentration ratio of alpha-particles to deuterons is $1.5\times10^{-4}$.

@@

\begin{align}
\vec{\vec{S}}_{s,n} &=
\begin{bmatrix}
\frac{n^2J_n^2}{z_s^2}v_\perp U_s & \frac{inJ_nJ_n'}{z_s}v_\perp U_s & \frac{n J_n^2}{z_s} v_\perp W_s \\
\frac{-inJ_nJ_n'}{z_s}v_\perp U_s & (J_n')^2 v_\perp U_s & -iJ_n J_n' v_\perp W_s \\
\frac{nJ_n^2}{z_s}v_\parallel U_s & iJ_nJ_n'v_\parallel U_s & n J_n^2 v_\parallel W_s \\
\end{bmatrix}\label{eq:stenor},\\
U_s &= \frac{\partial f_s}{\partial v_\perp} + \frac{k_\parallel}{\omega}  \left(v_\perp \frac{\partial f_s}{\partial v_\parallel} - v_\parallel \frac{\partial f_s}{\partial v_\perp}\right)\label{eq:uterm},\\
W_s &=\left(1- \frac{n\Omega_s}{\omega}\right)\frac{\partial f_s}{\partial v_\parallel} + \frac{n\Omega v_\parallel}{\omega v_\perp}\frac{\partial f_s}{\partial v_\perp}\label{eq:wterm},
\end{align}

@@

The growth rates of unstable modes for the case where the pitch-angle cosine
value is zero ($u_\parallel = 0$ in Eq. 1) are plotted in Fig.
\ref{fig:2D_F12_zeropitch}(a) against the associated real frequency (only modes
with growth rates in excess of the solver tolerance are shown i.e. $\Im(\omega)
> 10^{-5} \Omega_i$). In this panel shading indicates the parallel wavenumber in
units of $\Omega_i/V_A$. Fig. \ref{fig:2D_F12_zeropitch}(b) plots the same data in the $(k_\parallel,\omega)$ plane where shading instead indicates growth rates. The simplicity of the consecutive unstable cyclotron harmonics visible in panel (a) betrays the complexity of the structure in the $(k_\parallel, \omega)$ plane. Integration over $k_\parallel$ projects the solutions onto the frequency axis, which loses the structure and produces narrow harmonics. Fig. \ref{fig:2D_imag_zeropitch} plots shows the same data as Fig. \ref{fig:2D_F12_zeropitch} with damped and normal modes in addition, except shading indicates the growth rates in $(k_\parallel, k_\perp)$ space. Overlaid contours indicate the real frequency in units of ion cyclotron harmonics. Only narrow regions around the cyclotron harmonics are unstable. Overlaid straight traces indicate propagation angle. No Doppler shift is present due to the lack of parallel drift in the alpha-particle ring-beam. The growth rate increases approximately linearly with cyclotron harmonic number, which is consistent with the literature\cite{Dendy1994}. Together these figures serves to introduce a 2D view (growth rates in $(k_\parallel, k_\perp)$ space) of traditionally 1D data (growth rates against frequency). This will become important when a non-zero $u_\parallel$ is taken into account next.

@@

Fig. \ref{fig:2D_F12_physicalpitch} plots solutions for the physical pitch-angle
case ($u_\parallel = -0.64 u_0$) in the same manner as Fig.
\ref{fig:2D_F12_zeropitch}. There are several points worth noting. First, the
growth rate is no longer linear with cyclotron harmonic number. Second,
harmonics are broader due to the non-zero parallel drift that gives rise to
strong Doppler shift with increasing $|k_\parallel|$. Third, doublet splitting
of the $10^{th}$ harmonic is observed: see Fig. 2 of Cottrell 1993 for a single
signal showing multiple doublet splittings; Fig. 3 of Dendy et al 1994 (also
Fig. 9 Dendy et al 1995) shows slices at constant propagation angle that when
all superimposed point towards doublet splitting, albeit under-resolved in
wavenumber space. The following describes how these three features are
connected. The doublet splitting arises due to the dip in growth rate where
$k_\parallel$ goes through zero combined with the change in sign of the Doppler
shift of the real frequency as $k_\parallel$ changes sign. The effect is
exaggerated for higher harmonics where the growth rate diminishes to zero for
perpendicular propagation; see e.g. the $12^{th}$ harmonic in Fig.
\ref{fig:2D_imag_physicalpitch} near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $\Omega_{i}$ is the ion cyclotron
frequency and $V_A$ is the Alfv{\'e}n speed. Other possible mechanisms for
doublet splitting could be grad-B or curvature
drifts\cite{Cottrell1993,Fulop1998}. Qualitatively, Fig. \ref{fig:2D_F12_physicalpitch}(a) resembles Fig. 2 of Cottrell 1993 in that the signal below $\sim 7$ cyclotron harmonics is lower and flatter with well defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change larger and the harmonics are less easily identifiable. These features can only be recreated with a 2D calculation for these distributions functions. Note that Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes more clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is characterised by narrow regions of instability near Doppler shifted cyclotron harmonics that are stabilised when $k_\parallel=0$ but unstable at larger negative parallel wavenumbers. The higher frequency behaviour in the region $\omega\geq 7\Omega_i$ is dominated by broader regions of instability in frequency and wavenumber near to Doppler shifted cyclotron harmonics, which have a complex structure in wavenumber space. Modes propagating exactly perpendicular become stable in the region $k_\perp > 13 \Omega_i/V_A$ whilst obliquely propagating modes remain unstable; see Fig. 9 of Dendy et al 1995, which plots the analytic expression for the growth rate of the magnetoacoustic cylcotron instability, i.e. the instability thought responsible for ICE, at 5 slices of propagation angle. The Doppler shift of the growing modes can reach $\approx \Omega_i$ for large propagation angles at lower frequencies and also for moderately oblique angles at higher frequencies; see, for example, the mode associated with the $15^{th}$ cyclotron harmonic at, $(k_\parallel \simeq -\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 14 \Omega_i/V_A$, i.e. a propagation angle of $93^\circ$). Furthermore, the breadth of the harmonics can be attributed to the contributions from multiple propagation angles even though the lower harmonics are stable for forward travelling and perpendicular propagation angles.

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig3.pdf}
    \caption{(colour online) The imaginary component of the frequencies for the case with parallel drift, $u_{\parallel} = -0.64 u_0 \simeq -1.085 V_A$. Panel (a) plots growth rate against frequency where shading indicates parallel wavenumber in units of $\Omega_i/V_A$, and (b) plots the solutions in the $k_\parallel, \omega$ plane where shading indicates growth rate. Note that the Doppler shift associated with the non-zero parallel drift speed broadens the unstable peaks in frequency. The 10th cyclotron harmonic is split into a doublet, or has two peaks because both the growth rate and real frequency vary with wavenumber. Compare with Fig. 2 of Cottrel et al 1993.}
    \label{fig:2D_F12_physicalpitch}
\end{figure}

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig4.pdf}
    \caption{(colour online) As in Fig. \ref{fig:2D_imag_zeropitch}, except for a ring-beam of alpha particles, with a physical pitch-angle value resulting in $u_{\parallel} = -0.64 u_0 \simeq -1.085 V_A$. The non-zero parallel drift speed causes up-down asymmetry; this figure for a pitch-angle cosine of the opposite sign, $u_{\parallel} = 0.64 u_0$, is identical expect flipped about $k_\parallel=0$.}
    \label{fig:2D_imag_physicalpitch}
\end{figure}

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig5.pdf}
    \caption{(colour online) As in Fig. \ref{fig:2D_imag_zeropitch}, except that the shading indicates the real part of the smoothed solution normalised by the high frequency expression for the fast Alfv{\'e}n wave set out by Eq. 25 of Ref. \cite{Dendy1994}, which is also Eq. 11 in Ref. \cite{McClements1996}. The differences in real frequency for the two sets of calculations, physical pitch-angle cosine of $-0.64$ and zero, are negligible.}
    \label{fig:2D_real_zeropitch}
\end{figure}

>>>>

Solutions to the linearised Maxwell-Vlasov system of equations are found in
$(k_\parallel, k_\bot)$ space for a ring-beam distribution of energetic minority
alpha-particles and Maxwellian electrons and background thermal majority
deuteron ions. Calculations are made for alpha-particles with ring-beams of two
pitch-angle cosine values: $u_\parallel=-0.646 u_0 \approx
-\cos(\arcsin(\sqrt{R_0 \sin(\psi_0)/R_1})) \simeq -1.1 V_A$, which is
associated with the ion-cyclotron emission observed on JET Pulse No. 26148 for
which $R_0=3.46m$ and
$\psi_0=55.2^\circ$ are the major radius and angle pitch angle at birth, and
$R_1=4m$ is the major radius of the emission region, see
Figs. 3 and 14 and associated text in Ref. \cite{Cottrell1993}; and
$u_\parallel=0 u_0$ representing a ring distribution with no parallel drift, see
Fig. 1(a) of Ref \cite{Cook2013}. Here $u_0=\sqrt{u_\parallel^2 + u_\perp^2}$ is the speed of an alpha-particle with an energy of 3.6 MeV. The complex frequency solutions, $\omega$, are found corresponding to the fast Alfv{\'e}n branch on a grid of $(k_\parallel, k_\bot)$.

The plasma parameters approximate those found in the region indicated in Fig. 3
of Ref. \cite{Cottrell1993}, which are associated with the continuous trace
signal of Fig. 2 (ibid). The electrons and deuterons, both modelled as
Maxwellians, have temperatures of 1 keV and the electrons have a number density
of $1.7\times 10^{19}$ $\mathrm{m^{-3}}$. The magnetic field is $2.07$
$\mathrm{T}$, which corresponds to a deuterium cyclotron frequency of $17MHz$.
The ring-beam distribution function, Eq. \ref{eq:alphaf}, is a separable function of $v_\parallel$, i.e. the beam given by Eq. 1, and $v_\perp$, i.e. the ring given by Eq. 2. The component parallel (perpendicular) to the magnetic field is a drifting Maxwellian with thermal speed $v_{\parallel, th}$ ($v_{\perp, th}$) and drift velocity $u_\parallel$ ($u_\perp$). The thermal spread of the alpha particles' distribution function $v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The concentration ratio of alpha-particles to deuterons is $1.5\times10^{-4}$.

@@

\begin{align}
\vec{\vec{S}}_{s,n} &=
\begin{bmatrix}
\frac{n^2J_n^2}{z_s^2}v_\perp U_s & \frac{inJ_nJ_n'}{z_s}v_\perp U_s & \frac{n
  J_n^2}{z_s} v_\perp W_{s,n} \\
\frac{-inJ_nJ_n'}{z_s}v_\perp U_s & (J_n')^2 v_\perp U_s & -iJ_n J_n' v_\perp
  W_{s,n} \\
\frac{nJ_n^2}{z_s}v_\parallel U_s & iJ_nJ_n'v_\parallel U_s & n J_n^2
  v_\parallel W_{s,n} \\
\end{bmatrix}\label{eq:stenor},\\
U_s &= \frac{\partial f_s}{\partial v_\perp} + \frac{k_\parallel}{\omega}  \left(v_\perp \frac{\partial f_s}{\partial v_\parallel} - v_\parallel \frac{\partial f_s}{\partial v_\perp}\right)\label{eq:uterm},\\
  W_{s,n} &=\left(1- \frac{n\Omega_s}{\omega}\right)\frac{\partial f_s}{\partial v_\parallel} + \frac{n\Omega v_\parallel}{\omega v_\perp}\frac{\partial f_s}{\partial v_\perp}\label{eq:wterm},
\end{align}

@@

The growth rates of unstable modes for the case where the pitch-angle cosine
value is zero ($u_\parallel = 0$ in Eq. 1) are plotted in Fig.
\ref{fig:2D_F12_zeropitch}(a) against the associated real frequency (only modes
with growth rates in excess of the solver tolerance are shown i.e. $\Im(\omega)
> 10^{-5} \Omega_i$). In this panel shading indicates parallel wavenumber in
units of $\Omega_i/V_A$. Fig. \ref{fig:2D_F12_zeropitch}(b) plots the same data in the $(k_\parallel,\omega)$ plane where shading instead indicates growth rates. The simplicity of the consecutive unstable cyclotron harmonics visible in panel (a) betrays the complexity of the structure in the $(k_\parallel, \omega)$ plane. Integration over $k_\parallel$ projects the solutions onto the frequency axis, which loses the structure and produces narrow harmonics. Fig. \ref{fig:2D_imag_zeropitch} plots shows the same data as Fig. \ref{fig:2D_F12_zeropitch} with damped and normal modes in addition, except shading indicates the growth rates in $(k_\parallel, k_\perp)$ space. Overlaid contours indicate the real frequency in units of ion cyclotron harmonics. Only narrow regions around the cyclotron harmonics are unstable. Overlaid straight traces indicate propagation angle. No Doppler shift is present due to the lack of parallel drift in the alpha-particle ring-beam. The growth rate increases approximately linearly with cyclotron harmonic number, which is consistent with the literature\cite{Dendy1994}. Together these figures serves to introduce a 2D view (growth rates in $(k_\parallel, k_\perp)$ space) of traditionally 1D data (growth rates against frequency). This will become important when a non-zero $u_\parallel$ is taken into account next.

@@

Fig. \ref{fig:2D_F12_physicalpitch} plots solutions for the physical pitch-angle
case ($u_\parallel = -0.646 u_0$) in the same manner as Fig.
\ref{fig:2D_F12_zeropitch}. There are several points worth noting. First, the
growth rate is no longer linear with cyclotron harmonic number. Second,
harmonics are broader due to the non-zero parallel drift that gives rise to
strong Doppler shift with increasing $|k_\parallel|$. Third, doublet splitting
of the $11^{th}$ harmonic is observed: see Fig. 2 of Cottrell 1993 for a single
signal showing multiple doublet splittings; Fig. 3 of Dendy et al 1994 (also
Fig. 9 Dendy et al 1995) shows slices at constant propagation angle that when
all superimposed point towards doublet splitting, albeit under-resolved in
wavenumber space. The following describes how these three features are
connected. The doublet splitting arises due to the dip in growth rate where
$k_\parallel$ goes through zero combined with the change in sign of the Doppler
shift of the real frequency as $k_\parallel$ changes sign. The effect is
exaggerated for higher harmonics where the growth rate diminishes to zero for
perpendicular propagation; see e.g. the $12^{th}$ harmonic in Fig.
\ref{fig:2D_imag_physicalpitch} near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $\Omega_{i}$ is the ion cyclotron
frequency and $V_A$ is the Alfv{\'e}n speed. Other possible mechanisms for
doublet splitting could be grad-B or curvature
drifts\cite{Cottrell1993,Fulop1998}. Qualitatively, Fig. \ref{fig:2D_F12_physicalpitch}(a) resembles Fig. 2 of Cottrell 1993 in that the signal below $\sim 7$ cyclotron harmonics is lower and flatter with well defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change larger and the harmonics are less easily identifiable. These features can only be recreated with a 2D calculation for these distributions functions. Note that Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes more clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is characterised by narrow regions of instability near Doppler shifted cyclotron harmonics that are stabilised when $k_\parallel=0$ but unstable at larger negative parallel wavenumbers. The higher frequency behaviour in the region $\omega\geq 7\Omega_i$ is dominated by broader regions of instability in frequency and wavenumber near to Doppler shifted cyclotron harmonics, which have a complex structure in wavenumber space. Modes propagating exactly perpendicular become stable in the region $k_\perp > 13 \Omega_i/V_A$ whilst obliquely propagating modes remain unstable; see Fig. 9 of Dendy et al 1995, which plots the analytic expression for the growth rate of the magnetoacoustic cylcotron instability, i.e. the instability thought responsible for ICE, at 5 slices of propagation angle. The Doppler shift of the growing modes can reach $\approx \Omega_i$ for large propagation angles at lower frequencies and also for moderately oblique angles at higher frequencies; see, for example, the mode associated with the $15^{th}$ cyclotron harmonic at, $(k_\parallel \simeq -\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 14 \Omega_i/V_A$, i.e. a propagation angle of $93^\circ$). Furthermore, the breadth of the harmonics can be attributed to the contributions from multiple propagation angles even though the lower harmonics are stable for forward travelling and perpendicular propagation angles.

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig3.pdf}
    \caption{(colour online) The imaginary component of the frequencies for the
    case with parallel drift, $u_{\parallel} = -0.646 u_0 \simeq -1.1 V_A$.
    Panel (a) plots growth rate against frequency where shading indicates
    parallel wavenumber in units of $\Omega_i/V_A$, and (b) plots the solutions
    in the $k_\parallel, \omega$ plane where shading indicates growth rate. Note
    that the Doppler shift associated with the non-zero parallel drift speed
    broadens the unstable peaks in frequency. The $10^{th}$ and $11^{th}$ cyclotron
    harmonics are split into doublets, i.e. they have two peaks because both
    the growth rate and real frequency vary with wavenumber. Compare with
    Fig. 2 of Cottrel et al 1993.}
    \label{fig:2D_F12_physicalpitch}
\end{figure}

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig4.pdf}
    \caption{(colour online) As in Fig. \ref{fig:2D_imag_zeropitch}, except for a ring-beam of alpha particles, with a physical pitch-angle value resulting in $u_{\parallel} = -0.646 u_0 \simeq -1.1 V_A$. The non-zero parallel drift speed causes up-down asymmetry; this figure for a pitch-angle cosine of the opposite sign, $u_{\parallel} = 0.646 u_0$, is identical expect flipped about $k_\parallel=0$.}
    \label{fig:2D_imag_physicalpitch}
\end{figure}

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig5.pdf}
    \caption{(colour online) As in Fig. \ref{fig:2D_imag_zeropitch}, except that the shading indicates the real part of the smoothed solution normalised by the high frequency expression for the fast Alfv{\'e}n wave set out by Eq. 25 of Ref. \cite{Dendy1994}, which is also Eq. 11 in Ref. \cite{McClements1996}. The differences in real frequency for the two sets of calculations, physical pitch-angle cosine of $-0.646$ and zero, are negligible.}
    \label{fig:2D_real_zeropitch}
\end{figure}
